### 2025-03-01 sports-cognitive-psychology.qmd
20250228 - plot tweaks

# labels on the plot
text_df <- tibble(
  label = c(
    "NFL coaches<br>in <span style='color:#00BFC4'>**2021**</span>",
    "NFL coaches<br>in <span style='color:#F8766D'>**2017**</span>"),
  x = c(6, 8.2),
  y = c(90, 37),
  angle = c(10, 10),
  color = c("black", "black")
)

nfl_4thdown %>%
  filter(
    vegas_wp > .2,
    between(go_boost, -10, 10),
    season %in% c(2017, 2021)) %>%
  ggplot(
    aes(
      x = go_boost,
      y = go,
      color = as.factor(season))) + 
  geom_vline(xintercept = 0) +
  stat_smooth(
    method = "gam",
    method.args = list(gamma = 1),
    formula = y ~ s(x, bs = "cs", k = 10),
    show.legend = FALSE,
    se = FALSE,
    linewidth = 4) +
  # this is just to get the plot to draw the full 0 to 100 range
  geom_hline(
    yintercept = 100,
    alpha = 0) +
  geom_hline(
    yintercept = 0,
    alpha = 0) +
  ggtext::geom_richtext(
    data = text_df,
    aes(
      x,
      y,
      label = label,
      angle = angle),
    color = "black",
    fill = NA,
    label.color = NA,
    size = 5) + 
  theme_classic() +
  labs(
    x = "Gain in win probability by going for it",
    y = "Go-for-it percentage",
    subtitle = "4th down decisions in 2021 versus 2017, win prob. > 20%",
    title = glue::glue("How <span style='color:red'>math</span> is changing football")) +
  theme(
    legend.position = "none",
    plot.title = element_markdown(
      size = 22,
      hjust = 0.5),
    plot.subtitle = element_markdown(
      size = 14,
      hjust = 0.5),
    axis.title.x = element_text(
      size = 14,
      face = "bold"),
    axis.title.y = element_text(
      size = 14,
      face = "bold",
      angle = 0, # horizontal y-axis title
      vjust = 0.5)
  ) +
  scale_y_continuous(
    breaks = scales::pretty_breaks(n = 4),
    limits = c(0, 100),
    expand = c(0, 0)) +
  scale_x_continuous(
    breaks = scales::pretty_breaks(n = 10),
    limits = c(-10, 10),
    expand = c(0, 0)) +
  annotate(
    "text",
    x = -1.2,
    y = 70,
    label = "Should\nkick",
    color = "black",
    size = 5) +
  annotate(
    "text",
    x = 1.2,
    y = 70,
    label = "Should\ngo for it",
    color = "black",
    size = 5) +
  annotate(
    "segment",
    x = -.1,
    y = 80,
    xend = -2,
    yend = 80,
    arrow = arrow(length = unit(0.05, "npc")),
    color = "black",
    linewidth = 2) +
  annotate(
    "segment",
    x = .1,
    y = 80,
    xend = 2,
    yend = 80,
    arrow = arrow(length = unit(0.05, "npc")),
    color = "black",
    linewidth = 2)
```

>>>>

# labels on the plot
text_df <- tibble(
  label = c(
    "NFL coaches<br>in <span style='color:#00BFC4'>**2021**</span>",
    "NFL coaches<br>in <span style='color:#F8766D'>**2017**</span>"),
  x = c(5, 7),
  y = c(90, 20),
  angle = c(10, 10),
  color = c("black", "black")
)

nfl_4thdown %>%
  filter(
    vegas_wp > .2,
    between(go_boost, -10, 10),
    season %in% c(2017, 2021)) %>%
  ggplot(
    aes(
      x = go_boost,
      y = go,
      color = as.factor(season))) + 
  geom_vline(xintercept = 0) +
  stat_smooth(
    method = "gam",
    method.args = list(gamma = 1),
    formula = y ~ s(x, bs = "cs", k = 10),
    show.legend = FALSE,
    se = FALSE,
    linewidth = 4) +
  # this is just to get the plot to draw the full 0 to 100 range
  geom_hline(
    yintercept = 100,
    alpha = 0) +
  geom_hline(
    yintercept = 0,
    alpha = 0) +
  ggtext::geom_richtext(
    data = text_df,
    aes(
      x,
      y,
      label = label,
      angle = angle),
    color = "black",
    fill = NA,
    label.color = NA,
    size = 5) + 
  theme_classic() +
  labs(
    x = "Gain in win probability by going for it",
    y = "Go-for-it percentage",
    subtitle = "4th down decisions in 2021 versus 2017, win prob. > 20%",
    title = glue::glue("How <span style='color:red'>math</span> is changing football")) +
  theme(
    legend.position = "none",
    plot.title = element_markdown(
      size = 22,
      hjust = 0.5),
    plot.subtitle = element_markdown(
      size = 13.5,
      hjust = 0.5),
    axis.title.x = element_text(
      size = 14,
      face = "bold"),
    axis.title.y = element_text(
      size = 14,
      face = "bold",
      angle = 0, # horizontal y-axis title
      vjust = 0.5)
  ) +
  scale_y_continuous(
    breaks = scales::pretty_breaks(n = 4),
    limits = c(0, 100),
    expand = c(0, 0)) +
  scale_x_continuous(
    breaks = scales::pretty_breaks(n = 10),
    limits = c(-10, 10),
    expand = c(0, 0)) +
  annotate(
    "text",
    x = -1.5,
    y = 70,
    label = "Should\nkick",
    color = "black",
    size = 5) +
  annotate(
    "text",
    x = 1.75,
    y = 70,
    label = "Should\ngo for it",
    color = "black",
    size = 5) +
  annotate(
    "segment",
    x = -.1,
    y = 80,
    xend = -2,
    yend = 80,
    arrow = arrow(length = unit(0.05, "npc")),
    color = "black",
    linewidth = 2) +
  annotate(
    "segment",
    x = .1,
    y = 80,
    xend = 2,
    yend = 80,
    arrow = arrow(length = unit(0.05, "npc")),
    color = "black",
    linewidth = 2)
```
